Scattering and Pair Processes in General Relativistic Ray Tracing for Neutrinos

### 2016-10-18 paper.tex
Fix typos

With a ray tracing method we approach radiation transport from the perspective
of a single observer at a spacetime event $x_o^\alpha$.
Our goal is to compute the distribution function $f(x_o^\alpha;p_\beta)$,
or the amount of radiation with momentum $p_\beta$ impinging on $x_o^\alpha$.
To do so we trace a geodesic trajectory in the backwards direction
$-p_\beta$ to sample the incoming radiation along that line of sight.
By tracing a family of rays intersecting $x_o^\alpha$ we build up a
picture of the distribution function there.
And by sampling many observation points we construct a global picture
of $f(x^\alpha;p_\beta)$.

The ray tracing framework
is conceptually simple because it interprets the equation of radiation
transport (Eqn.~\ref{eqn:boltzmann})
as a one-dimensional ordinary differential equation,
it is numerically cheap because it confines computations of
$f(x_o^\alpha;p_\beta)$ to the past light-cone of $x_o^\alpha$,
and it easily extends to high resolution in energy and angle by simply
increasing the number of rays sampled.

@@

Greek tensor indices ($\alpha, \beta, ...$) range over all four coordinates,
whereas Latin indices ($i, j, ...$) range over the spatial coordinates 1--3.
We use naturalized units in which $\hbar=c=1$.

@@

Because $\ell_\beta$ is subject to two constraints
(normalization and orthogonality to the observer's velocity)
it has two degrees of freedom; we this make explicit by defining its
spatial components with respect to spherical polar angles
\begin{equation}
  \label{eqn:def_direction}
  \ell_\alpha :=
  q (s,\sin\upsilon_1\cos\upsilon_2,\sin\upsilon_1\sin\upsilon_2,\cos\upsilon_1),
\end{equation}
with $q$ and $s$ functions of $\upsilon_1$ and $\upsilon_2$.
Now our symbol for the distribution function,
$f(x^\alpha;\varepsilon,\upsilon_1,\upsilon_2)$,
makes manifest its seven independent arguments.

\subsection{Boltzmann Equation}
\label{ssec:boltzmann}
Neutrino radiation obeys the relativistic Boltzmann Equation
\todo{point out limit of QKEs}
\begin{equation}
  \label{eqn:boltzmann}
  \frac{d}{d\lambda}f(x^\alpha;\varepsilon,\upsilon_1,\upsilon_2) = C[f],
\end{equation}
where $\frac{d}{d\lambda}$ denotes a derivative with respect to the affine
parameter defining the neutrino momentum (Eqn.~\ref{eqn:geodesic_x})
and $C[f]$ is the source term arising from interactions with the medium.
The source term varies over phase space $(x^\alpha,p_\beta)$,
and depends locally on the distribution function
$f(x^\alpha;\varepsilon,\upsilon_1,\upsilon_2)$,
positon $x^\alpha$, and momentum $p_\beta$ of the neutrinos in question,
and nonlocally on the distribution function of this neutrino and its
antiparticle at different momenta, $f'$ and $\bar{f}'$.
For clarity we symbolize all of these dependencies with the shorthand $C[f]$.
The various neutrino interactions contributing to $C[f]$ are detailed in
App.~\ref{sec:source_terms}.

@@

\subsection{Neutrinosurface Approximation}
\label{ssec:neutrinosurface}
\todo{derive $\tau=2/3$ approx \cite[Chap.~4]{shu1991-book_radiation},
  and \cite{miln1921-radiative_equil_1}}
In the simplified configuration of a plane-parallel atmosphere, the
rendering equation reduces to 
\begin{equation}
\label{eqn:neutrinosurface}
  f(\lambda) = f^{\rm eq}(\lambda_e),
\end{equation}
where $\tau(\lambda,\lambda_e)=2/3$.
At an optical depth of $2/3$, the energy flux is equivalent to the
the blackbody value $\sigma T^4$, with $\sigma$, the Stefan-Boltzmann radiation
constant.
\todo{verify}

@@

We follow
\cite{brue1985-core_collapse} and \cite[Sec.~4]{shib2011-truncated_moment}
by separating these processes into four categories:
\begin{equation}
  \label{eqn:four_sources}
  C[f] \equiv C_{\rm AE} + C_{\rm ES} + C_{\rm IS} + C_{\rm PP},
\end{equation}
representing charged-current absorption and emission, elastic scattering,
inelastic scattering, and thermal pair annihilation and production. Each term
is computed by summing the weak interaction rates of the processes from the
relevant category given in Tab.~\ref{tab:neutrino_processes}.
These rates are computed most easily in the fluid frame of reference;
but since we use them in the form of an invariant emissivity and opacity
(Eqn.~\ref{eqn:boltzmann_linear}) they take the same numerical value in any
frame of reference and for any choice of fiducial observer $u^\alpha$.
\todo{confirm}

In practice we compute our rates using the open source neutrino interaction
library \lstinline{NuLib}~\footnote{\url{http://www.nulib.org}}.
We compile a table of sources defined over the four dimensions of
density, temperature, electron-fraction, and neutrino energy, and
interpolate quad-linearly to the points sampled along each ray.
\todo{derive and type up all sources in terms of $J$ and $H^\alpha$
  following \cite[Sec.~4]{shib2011-truncated_moment}}

\begin{table}[h]
  \caption{
    We analyze neutrino interaction processes in terms of these categories.
    $\nu$ without a label represents a neutrino or antineutrino of any flavor,
    $N$ represents a nucleon $n$ or $p$,
    ${}^ZA$ represents a nucleus with mass number $A$ and charge $Z$, and
    $\gamma$ represents a high-energy photon.
    A prime indicates a change in that particle's energy.
  }
  \label{tab:neutrino_processes}
  \begin{tabularx}{\columnwidth}{X X}
    \hline \hline
    absorption/emission
    & $\nu_e + n \leftrightarrow e^- + p$                          \\
    & $\bar{\nu}_e + p \leftrightarrow e^+ + n$                    \\
    & $\nu_e + {}^AZ \leftrightarrow e^- + {}^A(Z+1)$              \\
    & $\bar{\nu}_e + {}^AZ \leftrightarrow e^+ + {}^A(Z-1)$        \\
    \hline
    elastic scattering
    & $\nu + N \leftrightarrow \nu + N$                            \\
    & $\nu + {}^AZ \leftrightarrow \nu + {}^AZ$                    \\
    \hline
    inelastic scattering
    & $\nu + e^- \leftrightarrow \nu' + e^{-'}$                    \\
    & $\nu + e^+ \leftrightarrow \nu' + e^{+'}$                    \\
    \hline
    thermal pair processes
    & $\nu + \bar{\nu} \leftrightarrow e^{-} + e^{+}$              \\
    & $\nu + \bar{\nu} + N + N \leftrightarrow N' + N'$            \\
    & $\nu + \bar{\nu} \leftrightarrow \gamma$                     \\
    \hline \hline
  \end{tabularx}
\end{table}

\subsection{Absorption and emission via charged current}
\label{ssec:sources_ae}
At neutrino and thermal energies well below $m_{\mu}\sim100\,{\rm MeV}$
only charged current processes involving $\nu_e$ and $\bar{\nu}_e$ are allowed.
For each of these processes, $i$, we may write an emission and absorption
coefficient as a function of the interaction cross-section
(e.g. \cite[Eqn.~A5]{brue1985-core_collapse}):
\todo{explain Fermi-blocking}
\begin{align}
  \label{eqn:sources_ae_1}
  C_{{\rm AE},i}
  &= \mathscr{E}_{{\rm AE},i} (1-f) - \mathscr{K}_{{\rm AE},i} f \\
  \label{eqn:sources_ae_2}
  &= \mathscr{E}_{{\rm AE},i} - (\mathscr{E}_{{\rm AE},i} + \mathscr{K}_{{\rm AE},i}) f,
\end{align}
with
\begin{equation}
  \label{eqn:inv_emissivity}
  \mathscr{E}_{{\rm AE},i} = \frac{(2\pi)^3}{\varepsilon^2}\eta_{i},
\end{equation}
where $\eta$ is the traditional emissivity with dimension
$[\eta]={\rm energy}\,{\rm length}^{-3}\,{\rm time}^{-1}\,{\rm energy}^{-1}\,{\rm steradian}^{-1}$,
and
\begin{equation}
  \label{eqn:inv_emissivity}
  \mathscr{K}_{{\rm AE},i} = \varepsilon \chi_i,
\end{equation}
where $\chi$ is the traditional opacity with dimension
$[\chi]={\rm length}^{-1}$.

\subsection{Elastic scattering}
\label{ssec:sources_es}
Neutrino scattering on particles of mass much greater than $\varepsilon$
is essentially iso-energetic
(e.g. \cite[Eqn.~A8]{brue1985-core_collapse}).
\begin{equation}
  \label{eqn:sources_es}
  C_{{\rm ES},i}
  = \mathscr{K}_{{\rm ES},i} \Phi_{{\rm ES},i} - \mathscr{K}_{{\rm ES},i} f,
\end{equation}
with
\begin{align}
  \label{eqn:es_factor}
  \Phi_{{\rm ES},i} &\equiv \frac{\varepsilon^3}{\mathscr{K}_{{\rm ES},i}}
  \oint d\Omega' R(\ell_\alpha,\ell'_\beta)f(\ell'_\beta) \\
  \label{eqn:es_opacity}
  \mathscr{K}_{{\rm ES},i} &\equiv \varepsilon^3
  \oint d\Omega' R(\ell_\alpha,\ell'_\beta),
\end{align}
where $R(\ell_\alpha,\ell'_\beta)$ is the scattering kernel from direction
$\ell'_\beta$ to direction $\ell_\alpha$
having dimension $[R]={\rm energy}^{-1}$.

>>>>

With a ray tracing method we approach radiation transport from the perspective
of a single observer at a spacetime event $x_o^\alpha$.
Our goal is to compute the distribution function
$f^{\nu_\sigma}(x_o^\alpha;p_\beta)$, or the amount of neutrino radiation
with momentum $p_\beta$ impinging on $x_o^\alpha$;
$\nu_\sigma$ labels this distribution function as describing one of the
six species of neutrinos or antineutrinos.
To do so we trace a geodesic trajectory in the backwards direction
$-p_\beta$ to sample the incoming radiation along that line of sight.
By tracing a family of rays intersecting $x_o^\alpha$ we build up a
picture of the distribution function there.
And by sampling many observation points we construct a global picture
of $f^{\nu_\sigma}(x^\alpha;p_\beta)$.

The ray tracing framework
is conceptually simple because it interprets the equation of radiation
transport (Eqn.~\ref{eqn:boltzmann})
as a one-dimensional ordinary differential equation,
it is numerically cheap because it confines computations of
$f^{\nu_\sigma}(x_o^\alpha;p_\beta)$ to the past light-cone of $x_o^\alpha$,
and it easily extends to high resolution in energy and angle by simply
increasing the number of rays sampled.

@@

Greek tensor indices ($\alpha, \beta, ...$) range over all four coordinates,
whereas Latin indices ($i, j, ...$) range over the spatial coordinates 1--3.
We use naturalized units in which $\hbar=c=1$.
And for the remainder of this article we suppress the neutrino species label
$\nu_\sigma$ since the formulation is general to any species.

@@

Because $\ell_\beta$ is subject to two constraints
(normalization and orthogonality to the observer's velocity)
it has two degrees of freedom; we make this explicit by defining its
spatial components with respect to spherical polar angles
\begin{equation}
  \label{eqn:def_direction}
  \ell_\alpha :=
  q (s,\sin\upsilon_1\cos\upsilon_2,\sin\upsilon_1\sin\upsilon_2,\cos\upsilon_1),
\end{equation}
with $q$ and $s$ functions of $\upsilon_1$ and $\upsilon_2$.
Now our symbol for the distribution function,
$f(x^\alpha;\varepsilon,\upsilon_1,\upsilon_2)$,
makes manifest its seven independent arguments.

\subsection{Boltzmann Equation}
\label{ssec:boltzmann}
Neutrino radiation obeys the relativistic Boltzmann Equation
\todo{point out limit of QKEs}
\begin{equation}
  \label{eqn:boltzmann}
  \frac{d}{d\lambda}f(x^\alpha;\varepsilon,\upsilon_1,\upsilon_2) = C[f],
\end{equation}
where $\frac{d}{d\lambda}$ denotes a derivative with respect to the affine
parameter defining the neutrino momentum (Eqn.~\ref{eqn:geodesic_x})
and $C[f]$ is the source term arising from interactions with the medium.
The source term varies over phase space $(x^\alpha,p_\beta)$,
and depends locally on the distribution function $f$
and nonlocally on the distribution function of this neutrino and its
antiparticle at different momenta, $f'$ and $\bar{f}'$.
For clarity we symbolize all of these dependencies with the shorthand $C[f]$.
The various neutrino interactions contributing to $C[f]$ are detailed in
App.~\ref{sec:source_terms}.

@@

\subsection{Neutrinosurface Approximation}
\label{ssec:neutrinosurface}
\todo{derive $\tau=2/3$ approx \cite[Chap.~4]{shu1991-book_radiation},
  and \cite{miln1921-radiative_equil_1}}
In the simplified configuration of a plane-parallel atmosphere, the
rendering equation reduces to 
\begin{equation}
\label{eqn:neutrinosurface}
  f(\lambda) = f^{\rm eq}(\lambda_e),
\end{equation}
where $\tau(\lambda,\lambda_e)=2/3$.
At an optical depth of $2/3$, the energy flux is equivalent to the
blackbody value $\sigma T^4$, with $\sigma$, the Stefan-Boltzmann radiation
constant.
\todo{verify}

@@

We follow
\cite{brue1985-core_collapse} and \cite[Sec.~4]{shib2011-truncated_moment}
by separating these processes into four categories:
\begin{equation}
  \label{eqn:four_sources}
  C[f] \equiv C_{\rm AE} + C_{\rm SE} + C_{\rm SI} + C_{\rm TP},
\end{equation}
representing charged-current absorption and emission, elastic scattering,
inelastic scattering, and thermal pair annihilation and production. Each term
is computed by summing the weak interaction rates of the processes from the
relevant category given in Tab.~\ref{tab:neutrino_processes}.
These rates are computed most easily as traditional emissivities and opacities
in the fluid frame of reference;
but since we use them in the form of an invariant emissivity and opacity
(defined e.g. in \ref{eqn:inv_emissivity} and \ref{eqn:inv_opacity},
and obeying Eqn.~\ref{eqn:boltzmann_linear})
they take the same numerical value in any frame of reference
and are completely independent of our choice of fiducial observer $u^\alpha$.
\todo{confirm}

Each source term is unique to the neutrino or antineutrino species modeled,
and consists of a sum over all of the processes contributing to that
category of interaction. For example:
$C^{\nu_e}_{{\rm AE}} = \sum_i\,C^{\nu_e}_{{\rm AE},i}$,
where $i$ labels the absorption/emission processes involving $\nu_e$
in Tab.~\ref{tab:neutrino_processes}.
By contrast $C^{\nu_x}_{{\rm AE}} = 0$,
where $\nu_x$ stands for any of the four heavy-lepton neutrinos and
antineutrinos.

We compute our rates using the open source neutrino interaction
library \lstinline{NuLib}~\footnote{\url{http://www.nulib.org}}.
We compile a table of sources defined over the four dimensions of
density, temperature, electron-fraction, and neutrino energy, and
interpolate quad-linearly to the points sampled along each ray.
\todo{derive and type up all sources in terms of $J$ and $H^\alpha$
  following \cite[Sec.~4]{shib2011-truncated_moment}}

\begin{table}%[h]
  \caption{
    We analyze neutrino interaction processes in terms of these categories.
    $\nu$ without a label represents a neutrino or antineutrino of any flavor,
    $N$ represents a nucleon $n$ or $p$,
    ${}^ZA$ represents a nucleus with mass number $A$ and charge $Z$, and
    $\gamma$ represents a high-energy photon.
    A prime indicates a change in that particle's energy.
  }
  \label{tab:neutrino_processes}
  \begin{tabularx}{\columnwidth}{X X}
    \hline \hline
    absorption/emission
    & $\nu_e + n \leftrightarrow e^- + p$                          \\
    & $\bar{\nu}_e + p \leftrightarrow e^+ + n$                    \\
    & $\nu_e + {}^AZ \leftrightarrow e^- + {}^A(Z+1)$              \\
    & $\bar{\nu}_e + {}^AZ \leftrightarrow e^+ + {}^A(Z-1)$        \\
    \hline
    elastic scattering
    & $\nu + N \leftrightarrow \nu + N$                            \\
    & $\nu + {}^AZ \leftrightarrow \nu + {}^AZ$                    \\
    \hline
    inelastic scattering
    & $\nu + e^- \leftrightarrow \nu' + e^{-'}$                    \\
    & $\nu + e^+ \leftrightarrow \nu' + e^{+'}$                    \\
    \hline
    thermal pair processes
    & $\nu + \bar{\nu} \leftrightarrow e^{-} + e^{+}$              \\
    & $\nu + \bar{\nu} + N + N \leftrightarrow N' + N'$            \\
    & $\nu + \bar{\nu} \leftrightarrow \gamma$                     \\
    \hline \hline
  \end{tabularx}
\end{table}

\subsection{Absorption and emission via charged current}
\label{ssec:sources_ae}
At neutrino and thermal energies well below $m_{\mu}\sim100\,{\rm MeV}$
only charged current processes involving $\nu_e$ and $\bar{\nu}_e$ are allowed.
For each of these processes, $i$, we may write an emission and absorption
coefficient as a function of the interaction cross-section
(e.g. \cite[Eqn.~A5]{brue1985-core_collapse}):
\todo{explain Fermi-blocking}
\begin{align}
  \label{eqn:sources_ae_1}
  C_{{\rm AE},i}
  &= \mathscr{E}_{{\rm AE},i} (1-f) - \mathscr{K}_{{\rm AE},i} f \\
  \label{eqn:sources_ae_2}
  &= \mathscr{E}_{{\rm AE},i} - (\mathscr{E}_{{\rm AE},i} + \mathscr{K}_{{\rm AE},i}) f,
\end{align}
with
\begin{equation}
  \label{eqn:inv_emissivity}
  \mathscr{E}_{{\rm AE},i} = \frac{(2\pi)^3}{\varepsilon^2}\eta_{i},
\end{equation}
where $\eta$ is the traditional emissivity with dimension
$[\eta]={\rm energy}\,{\rm length}^{-3}\,{\rm time}^{-1}\,{\rm energy}^{-1}\,{\rm steradian}^{-1}$,
and
\begin{equation}
  \label{eqn:inv_opacity}
  \mathscr{K}_{{\rm AE},i} = \varepsilon \chi_i,
\end{equation}
where $\chi$ is the traditional opacity with dimension
$[\chi]={\rm length}^{-1}$.

\subsection{Elastic scattering}
\label{ssec:sources_es}
Neutrino scattering on particles of mass much greater than $\varepsilon$
is essentially iso-energetic
(e.g. \cite[Eqn.~A8]{brue1985-core_collapse}).
\begin{equation}
  \label{eqn:sources_es}
  C_{{\rm SE},i}
  = \mathscr{K}_{{\rm SE},i} \Phi_{{\rm SE},i} - \mathscr{K}_{{\rm SE},i} f,
\end{equation}
with
\begin{align}
  \label{eqn:es_factor}
  \Phi_{{\rm SE},i} &\equiv \frac{\varepsilon^3}{\mathscr{K}_{{\rm SE},i}}
  \oint d\Omega' R(\ell_\alpha,\ell'_\beta)f(\ell'_\beta) \\
  \label{eqn:es_opacity}
  \mathscr{K}_{{\rm SE},i} &\equiv \varepsilon^3
  \oint d\Omega' R(\ell_\alpha,\ell'_\beta),
\end{align}
where $R(\ell_\alpha,\ell'_\beta)$ is the scattering kernel from direction
$\ell'_\beta$ to direction $\ell_\alpha$
having dimension $[R]={\rm energy}^{-1}$.
